Predicting how climate change will affect how and where terrestrial mammals will move in British Columbia, Canada

### 2025-05-16 writing/appendix-b-gams.Rmd
added GLRT results, moved AIC stuff to app. B

---
title: \Large Predicting how climate change will affect how and where terrestrial mammals will move in British Columbia, Canada
subtitle: "Appendix B: Modeling the effects of temperature on movement behavior"
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: [braes, biol]
  - name: Chris H. Fleming
    institute: [ucf, smith]
  - name: Siobhan Darlington
    institute: [braes, biol]
  - name: Adam T. Ford
    institute: [braes, biol]
  - name: Karen E. Hodges
    institute: [braes, biol]
  - name: Kirk Safford
    institute: bcp
  - name: Robert Serrouya
    institute: [braes, biol, biodiv]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
  - bcp: BC Parks
  - biodiv: Wildlife Science Centre, Biodiversity Pathways, University of British Columbia Okanagan, Revelstoke, British Columbia, Canada.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'bc-mammals-temperature.bib'
csl: 'freshwater-biology.csl'
# csl: 'nature.csl'
fontsize: 12pt
indent: true
header-includes:
    - \renewcommand{\figurename}{Figure S\!\!} # for "Figure Ax.
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # for Delta symbol in table
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: true
---

@@

>>>>

---
title: \Large Predicting how climate change will affect how and where terrestrial mammals will move in British Columbia, Canada
subtitle: "Appendix B: Modeling the effects of temperature on movement behavior"
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: [braes, biol]
  - name: Chris H. Fleming
    institute: [ucf, smith]
  - name: Siobhan Darlington
    institute: [braes, biol]
  - name: Adam T. Ford
    institute: [braes, biol]
  - name: Karen E. Hodges
    institute: [braes, biol]
  - name: Kirk Safford
    institute: bcp
  - name: Robert Serrouya
    institute: [braes, biol, biodiv]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
  - bcp: BC Parks
  - biodiv: Wildlife Science Centre, Biodiversity Pathways, University of British Columbia Okanagan, Revelstoke, British Columbia, Canada.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'bc-mammals-temperature.bib'
csl: 'movement-ecology.csl'
# csl: 'nature.csl'
fontsize: 12pt
indent: true
header-includes:
    - \renewcommand{\figurename}{Figure B\!\!} # for "Fig. Bx.
    - \renewcommand{\tablename}{Table B\!\!} # for "Table. Bx.
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # for Delta symbol in table
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: true
---

@@

# Model selection

```{r temp-comparison}
setwd('..')
library('dplyr')
source('analysis/figures/default-ggplot-theme.R') #' for `SPECIES` object

temp_comp <- tibble(
  Model = c('$P(M)$', 'Speed', rep('HRSF', 7)),
  Species = c('All', 'All', sort(as.character(SPECIES))),
  DDE =
    c(10.72 - 10.23, 79.26 - 79.11, 9.16, 4.62, 18.6, 8.0, 5.30, 12.0, 17.7) %>%
    round(1) %>%
    paste0('%'),
  DAIC = c(5927, 1682, 6566, 4448, 813, 966, 5637, 342, 914) %>%
    round())

if(FALSE) {
  range(as.numeric(gsub('%', '', temp_comp$DDE)))
  range(as.numeric(gsub('%', '', temp_comp$DDE))[- c(1:2)])
  max(temp_comp$DAIC)
}

colnames(temp_comp)[colnames(temp_comp) == 'DAIC'] <- paste0('\U0394', 'AIC')
colnames(temp_comp)[colnames(temp_comp) == 'DDE'] <- paste0('\U0394', 'DE')

knitr::kable(temp_comp, digits = 1, format = 'markdown', caption = "Changes in deviance expained ($\\Delta$DE) and Akaike Information Criterion ($\\Delta$AIC) from including temperature in the models estimating probability of movement ($P(M)$) and speed when moving (Speed) as well as in species-level Hierarchical Resource Selection Functions (HRSF). Unlike conventional marginal AIC [@akaike_new_1974], The AIC values provided by the `mgcv` package for `R` are calculated using the models' estimated effective degrees of freedom and the Bayesian variance-covariance matrices, which avoids issues related to random effects or Restricted Maximum Likelihood (REML).")
```

## Analyses of Deviance

\noindent Here we produce approximate p-values for analyses of deviance (a form of generalized likelihood ratio tests) following the methods of Section 3.3 in @wood_generalized_2017. We use an $F$ test rather than a $\chi^2$ test for the Gamma models because the scale parameter for the Gamma distribution is estimated rather than known or fixed [as in the case of binomial or Poisson distributions -- see table 3.1 on page 104 of @wood_generalized_2017].

\singlespacing

\footnotesize

```{r, echo=TRUE}
# P(moving)
anova(readRDS('../models/binomial-gam-without-temperature.rds'),
      readRDS('../models/binomial-gam.rds'),
      test = 'Chisq')
```

```{r, echo=TRUE}
# speed when moving
anova(readRDS('../models/gamma-gam-without-temperature.rds'),
      readRDS('../models/gamma-gam.rds'),
      test = 'F')
```

<!-- \clearpage -->

```{r, echo=TRUE}
# HRSFs
anova(readRDS('../models/rsf-Canis lupus-2025-01-20.rds'),
      readRDS('../models/rsf-Canis lupus-no-temperature-2025-01-20.rds'),
      test = 'Chisq')
```

```{r, echo=TRUE}
anova(readRDS('../models/rsf-Cervus canadensis-2025-01-20.rds'),
      readRDS('../models/rsf-Cervus canadensis-no-temperature-2025-01-20.rds'),
      test = 'Chisq')
```

<!-- \clearpage -->

```{r, echo=TRUE}
anova(readRDS('../models/rsf-Oreamnos americanus-2025-01-20.rds'),
      readRDS('../models/rsf-Oreamnos americanus-no-temperature-2025-01-20.rds'),
      test = 'Chisq')
```

```{r, echo=TRUE}
anova(readRDS('../models/rsf-Puma concolor-2025-01-20.rds'),
      readRDS('../models/rsf-Puma concolor-no-temperature-2025-01-20.rds'),
      test = 'Chisq')
```

<!-- \clearpage -->

```{r, echo=TRUE}
anova(
  readRDS('../models/rsf-Rangifer tarandus boreal-2025-01-21.rds'),
  readRDS('../models/rsf-Rangifer tarandus boreal-no-temperature-2025-01-20.rds'),
  test = 'Chisq')
```

```{r, echo=TRUE}
anova(readRDS('../models/rsf-Rangifer tarandus southern mountain-2025-01-20.rds'),
      readRDS('../models/rsf-Rangifer tarandus southern mountain-no-temperature-2025-01-20.rds'),
      test = 'Chisq')
```

<!-- \clearpage -->

```{r, echo=TRUE}
anova(readRDS('../models/rsf-Ursus arctos horribilis-2025-01-20.rds'),
      readRDS('../models/rsf-Ursus arctos horribilis-no-temperature-2025-01-20.rds'),
      test = 'Chisq')
```

\normalsize

\doublespacing

# Refences

### 2025-05-16 writing/manuscript.Rmd
added GLRT results, moved AIC stuff to app. B

\noindent \textbf{Tables}: 3

\noindent \textbf{References}: 142 (updated on 2025-05-09) <!-- counted by using movement ecology csl -->

@@

This paper provides an analysis of the effects of air temperature on the movement of six terrestrial mammal species in British Columbia (BC), Canada (*Canis lupus*, *Cervus canadensis*, *Oreamnos americanus*, *Puma concolor*, boreal and southern mountain *Rangifer tarandus*, and *Ursus arctos horribilis*; Table \@ref(tab:data-table)). Using over 25 years of telemetry data thoughout a large spatial range of British Columbia and hourly air temperature data, we estimate how mammals altered their probability of movement, speed when moving, and habitat selection in response to temperature. We then pair the estimated responses with climate change projections to forecast the species' movement during the 21^st^ century under different climate-change scenarios [referred to as Shared Socioeconomic Pathways, SSPs -- see @riahi_shared_2017]. We disentangle the direct and indirect effects of temperature on the movement behaviour of terrestrial mammals and discuss the consequences of these changes at the individual, population, and landscape levels.

@@

Since `ctmm``s movement models assume stochastic but non-zero speeds (i.e., a single, stochastic moving state), data-informed speeds needed to be corrected so that the minimum instantaneous speed could be 0. We performed this correction by subtracting each model's mean speed while assuming speeds were $\chi^2$-distributed. The function we used is available on GitHub at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R). To avoid artifacts due to excessively small, non-zero speeds, we determined whether an animal was moving or not using the inflection point of each species' histogram of detrended speed estimates, as $k$-nearest neighbors did not discriminate between states well (Fig. S3).

@@

\noindent The following sections detail the statistical framework and modelling we used to estimate the effect on temperature on mammals' movement. To assess the importance of including temperature as an explicit covariate (rather than including its effects with time of day and day of year), we fit models with and without smooth effects of temperature and compared the deviance explained, model predictions, and AIC values of the two sets of models. We provide the relevant Directed Acyclical Graphs in the Appendix (Figs. S1-S2).

@@

\noindent In this study, we separate the effects of temperature on mammalian movement into indirect and direct effects. We call "indirect" all effects that can be intervened on by altering an animal's habitat [its geographic space *sensu* @matthiopoulos_defining_2022], and we refer to all other effects as "direct". For example, an increase in temperature may push an animal to prefer higher forest cover in search of shade, and conservationists can thus mitigate these indirect effects of temperature on movement behaviour by increasing the availability of forested areas, since the effect of temperature on movement is conditional on forest cover availability. In contrast, the animal may avoid moving if temperatures are above 35$^\circ$C, irrespective of its geographic and environmental space [*sensu* @matthiopoulos_defining_2022], so it is not possible to mitigate the effects of temperature on movement behaviour (other than changing temperature directly). We provide the relevant Directed Acyclical Graphs in the Appendix (Figs. S1-S2).

@@

Smooth effects of percent forest cover, elevation, and distance to water accounted for the species-level average selection strength for each resource. A random effect for each individual animal (`s(animal, bs = 're')`) corrected for uneven sampling across individuals, while factor smooth interaction terms (`bs = 'fs'`) accounted for individual-level resource selection (i.e., individual-level deviations from the species-level average). Tensor interaction product terms (`ti()`) of the three resources and temperature estimated the change in resource selection at different temperatures. Finally, we included marginal smooth terms of temperature to account for species- and individual-level sampling biases over temperature (e.g., sampling more during warm periods). Detections were weighted proportionally to their degree of independence from other temporally proximate detections [`weights = weight` -- see Appendix B and @alston_mitigating_2022], while quadrature points had a weight of 1. Quadrature points were obtained using all raster cells in the 99.9% AKDE percentile. The number of quadrature locations greatly outnumbered the number of observed locations (Fig. S13), especially after accounting for weighting based on the degree of autocorrelation and the number of home range crossings (Fig. S14).

@@

\noindent Data were relatively scarce for temperatures lower than $-20^\circ$C or above $20^\circ$C (Fig. S4). Species differed in overall mean probabilities of movement (range: 0.05 -- 0.3), mean speed when moving (range: 0.42 -- 2.7 km/day), and mean distance traveled (range: 0.04 -- 0.6 km/day; Table \@ref(tab:gam-intercepts)). Grizzly bears tended to move least often (probability of moving $P(M) \approx 0.05$), while wolves and pumas moved most often ($P(M) \gtrapprox 0.23$). When moving, mountain goats and southern mountain caribou moved the slowest ($\mathbb E(S|M) \approx 0.4$ m/s), while wolves had the highest mean speed when moving ($\mathbb E(S|M) \approx 2.7$ m/s). Consequently, wolves traveled, on average, 50 km/day -- 2.5 to 16.7 times further than other mammals.

@@

Including temperature in the models always increased within-sample prediction, but the change was not always substantial (deviance explained increased between +0.2% and +18.6%, with all HRSFs having an increase $\ge +4.6$%), and there was relatively strong agreement between models with and without temperature (Figs. S5 and S13), but including temperature resulted in substantially lower AIC values for all models, especially HRSFs (all $\Delta$AIC $\ge$ 342; Table \@ref(tab:temp-comparison)). Although this can be interpreted as evidence for a better fit, it is important to note that the addition of terms generally decreases AIC when sample sizes are large, even if the more complex model is not closer to the true model [see section 1.8.6, pages 52-54 of @wood_generalized_2017].

```{r temp-comparison}
setwd('..')
library('dplyr')
source('analysis/figures/default-ggplot-theme.R') #' for `SPECIES` object

temp_comp <- tibble(
  Model = c('$P(M)$', 'Speed', rep('HRSF', 7)),
  Species = c('All', 'All', sort(as.character(SPECIES))),
  DDE = 
    c(10.72 - 10.23, 79.26 - 79.11, 9.16, 4.62, 18.6, 8.0, 5.30, 12.0, 17.7) %>%
    round(1) %>%
    paste0('%'),
  DAIC = c(5927, 1682, 6566, 4448, 813, 966, 5637, 342, 914) %>%
    round())

if(FALSE) {
  range(as.numeric(gsub('%', '', temp_comp$DDE)))
  range(as.numeric(gsub('%', '', temp_comp$DDE))[- c(1:2)])
  max(temp_comp$DAIC)
}

colnames(temp_comp)[colnames(temp_comp) == 'DAIC'] <- paste0('\U0394', 'AIC')
colnames(temp_comp)[colnames(temp_comp) == 'DDE'] <- paste0('\U0394', 'DE')

knitr::kable(temp_comp, digits = 1, format = 'markdown', caption = "Changes in deviance expained ($\\Delta$DE) and Akaike Information Criterion ($\\Delta$AIC) from including temperature in the models estimating probability of movement (P(M)) and speed when moving (Speed) as well as in species-level Hierarchical Resource Selection Functions (HRSF). Unlike conventional marginal AIC [@akaike_new_1974], The AIC values provided by the `mgcv` package for `R` are calculated using the models' estimated effective degrees of freedom and the Bayesian variance-covariance matrices, which avoids issues related to random effects or Restricted Maximum Likelihood (REML).")
```

@@

\noindent The effects of temperature on movement rates varied in both direction and magnitude across species, even after accounting for differences in daily and seasonal activity (e.g., sleeping, migration, hibernation; see Figs. S6-S8). Smooth interaction terms were well-behaved and indicated clear shifts in activity over time of day and day of year for all species. The models explained reasonably high proportions of the deviance (10.7% for the binomial model and 79.3% for the Gamma model) and had good in-sample prediction (Fig. S9). All species adapted their daily and seasonal movement behaviour to changes in temperature (Fig. \@ref(fig:dist)). For example, wolves, elk, pumas, and boreal caribou reduced their mid-day and summer movement rates when temperatures were above 20$^\circ$C. However, uncertainties around the estimated effects were often higher at extreme temperatures due to data scarcity. All species' estimated probability of moving decreased with sampling intervals approximately above 1 hour, and most species' estimated speed when moving decreased monotonically with sampling interval, although the estimated trends were highly uncertain for some species (Fig. S10).

@@

Species' relative selection strength (RSS) was generally strongest for elevation and weakest for forest cover, but species' RSS for resources depended on temperature. Changes in RSS with temperature were also strongest for elevation and generally weakest distance from water, but there were no common trends across all species for any of the three resources. All species, with the exception of pumas, exhibited a clear temperature-dependent shift in their preference for forest cover. At higher temperatures, wolves, mountain goats, and grizzly bears became less selective for forest cover, while elk and caribou shifted towards more intermediate forest cover without much of a change in preference width. All species shifted elevationally with temperature, although boreal caribou did not exhibit as strong of a change. As temperatures rose, elk, mountain goats, and pumas increased in elevation, while boreal wolves, southern mountain caribou, and grizzly bears decreased in elevation. Most species generally remained within 5 km of water, and temperature did not affect their selection strength as much as for the other two resources. Again, estimated RSS values were generally most uncertain at extreme temperatures (Fig. S13).

>>>>

\noindent \textbf{Tables}: 2

\noindent \textbf{References}: 141 (updated on 2025-05-16) <!-- counted by using movement ecology csl -->

@@

This paper provides an analysis of the effects of air temperature on the movement of six terrestrial mammal species in British Columbia (BC), Canada (wolves: *Canis lupus*, elk: *Cervus canadensis*, mountain goats: *Oreamnos americanus*, pumas: *Puma concolor*, boreal and southern mountain caribou: *Rangifer tarandus*, and grizly bears: *Ursus arctos horribilis*; Table \@ref(tab:data-table)). Using over 25 years of telemetry data thoughout a large spatial range of British Columbia and hourly air temperature data, we estimate how mammals altered their probability of movement, speed when moving, and habitat selection in response to temperature. We then pair the estimated responses with climate change projections to forecast the species' movement during the 21^st^ century under different climate-change scenarios [referred to as Shared Socioeconomic Pathways, SSPs -- see @riahi_shared_2017]. We disentangle the direct and indirect effects of temperature on the movement behaviour of terrestrial mammals and discuss the consequences of these changes at the individual, population, and landscape levels.

@@

Since `ctmm``s movement models assume stochastic but non-zero speeds (i.e., a single, stochastic moving state), data-informed speeds needed to be corrected so that the minimum instantaneous speed could be 0. We performed this correction by subtracting each model's mean speed while assuming speeds were $\chi^2$-distributed. The function we used is available on GitHub at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R). To avoid artifacts due to excessively small, non-zero speeds, we determined whether an animal was moving or not using the inflection point of each species' histogram of detrended speed estimates, as $k$-nearest neighbors did not discriminate between states well (Fig. B3).

@@

\noindent The following sections detail the statistical framework and modelling we used to estimate the effect on temperature on mammals' movement. To assess the importance of including temperature as an explicit covariate (rather than including its effects with time of day and day of year), we fit models with and without smooth effects of temperature and compared the fits of the two sets of models. We provide the relevant Directed Acyclical Graphs in the Appendix (Figs. B1-B2).

@@

\noindent In this study, we separate the effects of temperature on mammalian movement into indirect and direct effects. We call "indirect" all effects that can be intervened on by altering an animal's habitat [its geographic space *sensu* @matthiopoulos_defining_2022], and we refer to all other effects as "direct". For example, an increase in temperature may push an animal to prefer higher forest cover in search of shade, and conservationists can thus mitigate these indirect effects of temperature on movement behaviour by increasing the availability of forested areas, since the effect of temperature on movement is conditional on forest cover availability. In contrast, the animal may avoid moving if temperatures are above 35$^\circ$C, irrespective of its geographic and environmental space [*sensu* @matthiopoulos_defining_2022], so it is not possible to mitigate the effects of temperature on movement behaviour (other than changing temperature directly). We provide the relevant Directed Acyclical Graphs in the Appendix (Figs. B1-B2).

@@

Smooth effects of percent forest cover, elevation, and distance to water accounted for the species-level average selection strength for each resource. A random effect for each individual animal (`s(animal, bs = 're')`) corrected for uneven sampling across individuals, while factor smooth interaction terms (`bs = 'fs'`) accounted for individual-level resource selection (i.e., individual-level deviations from the species-level average). Tensor interaction product terms (`ti()`) of the three resources and temperature estimated the change in resource selection at different temperatures. Finally, we included marginal smooth terms of temperature to account for species- and individual-level sampling biases over temperature (e.g., sampling more during warm periods). Detections were weighted proportionally to their degree of independence from other temporally proximate detections [`weights = weight` -- see Appendix B and @alston_mitigating_2022], while quadrature points had a weight of 1. Quadrature points were obtained using all raster cells in the 99.9% AKDE percentile. The number of quadrature locations greatly outnumbered the number of observed locations (Fig. B13), especially after accounting for weighting based on the degree of autocorrelation and the number of home range crossings (Fig. B14).

@@

\noindent Data were relatively scarce for temperatures lower than $-20^\circ$C or above $20^\circ$C (Fig. B4). Species differed in overall mean probabilities of movement (range: 0.05 -- 0.3), mean speed when moving (range: 0.42 -- 2.7 km/day), and mean distance traveled (range: 0.04 -- 0.6 km/day; Table \@ref(tab:gam-intercepts)). Grizzly bears tended to move least often (probability of moving $P(M) \approx 0.05$), while wolves and pumas moved most often ($P(M) \gtrapprox 0.23$). When moving, mountain goats and southern mountain caribou moved the slowest ($\mathbb E(S|M) \approx 0.4$ m/s), while wolves had the highest mean speed when moving ($\mathbb E(S|M) \approx 2.7$ m/s). Consequently, wolves traveled, on average, 50 km/day -- 2.5 to 16.7 times further than other mammals.

@@

There was relatively strong agreement between models with and without temperature (Figs. B5 and S13), but including temperature always resulted in better fits. All analyses of deviance showed that temperature significantly affected movement behavior (all p-values $< 2.2\times10^{-16}$; all $\Delta$AIC $\ge$ 342; see Table B1 and the analyses of deviance in Appendix B). The only exception to this was the mountain goat HRSF, where the addition of temperature resulted in a simpler model (lower model degrees of freedom) that fit the data better (lower deviance), and hence no test was necessary to show that the inclusion of temperature produced a better fit.

@@

\noindent The effects of temperature on movement rates varied in both direction and magnitude across species, even after accounting for differences in daily and seasonal activity (e.g., sleeping, migration, hibernation; see Figs. B6-B8). Smooth interaction terms were well-behaved and indicated clear shifts in activity over time of day and day of year for all species. The models explained reasonably high proportions of the deviance (10.7% for the binomial model and 79.3% for the Gamma model) and had good in-sample prediction (Fig. B9). All species adapted their daily and seasonal movement behaviour to changes in temperature (Fig. \@ref(fig:dist)). For example, wolves, elk, pumas, and boreal caribou reduced their mid-day and summer movement rates when temperatures were above 20$^\circ$C. However, uncertainties around the estimated effects were often higher at extreme temperatures due to data scarcity. All species' estimated probability of moving decreased with sampling intervals approximately above 1 hour, and most species' estimated speed when moving decreased monotonically with sampling interval, although the estimated trends were highly uncertain for some species (Fig. B10).

@@

Species' relative selection strength (RSS) was generally strongest for elevation and weakest for forest cover, but species' RSS for resources depended on temperature. Changes in RSS with temperature were also strongest for elevation and generally weakest distance from water, but there were no common trends across all species for any of the three resources. All species, with the exception of pumas, exhibited a clear temperature-dependent shift in their preference for forest cover. At higher temperatures, wolves, mountain goats, and grizzly bears became less selective for forest cover, while elk and caribou shifted towards more intermediate forest cover without much of a change in preference width. All species shifted elevationally with temperature, although boreal caribou did not exhibit as strong of a change. As temperatures rose, elk, mountain goats, and pumas increased in elevation, while boreal wolves, southern mountain caribou, and grizzly bears decreased in elevation. Most species generally remained within 5 km of water, and temperature did not affect their selection strength as much as for the other two resources. Again, estimated RSS values were generally most uncertain at extreme temperatures (Fig. B13).
